---
title: "The Story of Javier"
subtitle: "Over a Century of Baby Name Trends in the United States"
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(tidyverse)
library(plotly)
```

```{r}
#| label: load-data
files <- fs::dir_ls("data", glob = "*yob*.txt")
baby_names <- read_csv(files, id = "file", col_names = c("name", "sex", "count")) |>
  mutate(year = parse_number(basename(file)))

javier <- baby_names |>
  filter(name == "Javier", sex == "M")
```

## Overview

The name **Javier** (the Spanish equivalent of Xavier, derived from the Basque name *Xabier*) has a rich history in the United States. Over the past century, more than **108,000** boys have been registered with the name in Social Security records.

::: {.callout-note appearance="simple"}
### Quick Facts at a Glance
- **First Recorded Appearance:** 1917 (5 babies)
- **Peak Year:** 2001 (2,637 babies)
- **Most Recent Year (2025):** 1,252 babies
- **Total Registered:** 108,463 babies
:::

## Interactive Trend (1917–2025)

Hover over the line to explore exact birth figures for any given year:

```{r}
#| label: javier-plot
p <- ggplot(javier, aes(x = year, y = count, group = 1, text = paste0("Year: ", year, "<br>Births: ", scales::comma(count)))) +
  geom_line(color = "#2c3e50", linewidth = 0.8) +
  geom_point(data = filter(javier, year == 2001), aes(x = year, y = count, group = 1), color = "#e74c3c", size = 2.5) +
  annotate("text", x = 2001, y = 2750, label = "Peak: 2001 (2,637)", color = "#e74c3c", fontface = "bold", size = 3.5) +
  labs(
    x = "Year",
    y = "Number of Births",
    caption = "Source: Social Security Administration"
  ) +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 8)) +
  theme_minimal(base_size = 13)

ggplotly(p, tooltip = "text")
```

## Historical Insights

- **The Early Years (1917–1950s):** Javier was relatively rare in official U.S. records, averaging under 100 births annually.
- **The Rapid Rise (1960s–1990s):** Reflecting demographic growth and greater cultural representation, the name saw steady growth, crossing 1,000 births per year by the mid-1970s and 2,000 by 1990.
- **The Peak (2001):** The name reached its highest popularity in 2001 with 2,637 births.
- **Recent Decades (2010s–2025):** Popularity has stabilized around 1,200 to 1,500 births annually, remaining an enduring favorite.

## Comparison: Javier vs. Micheal

How does Javier compare to **Micheal** over the past century?

```{r}
#| label: comparison-plot
comparison <- baby_names |>
  filter(name %in% c("Javier", "Micheal"), sex == "M")

p_comp <- ggplot(comparison, aes(x = year, y = count, color = name, group = name, text = paste0("Name: ", name, "<br>Year: ", year, "<br>Births: ", scales::comma(count)))) +
  geom_line(linewidth = 0.9) +
  labs(
    x = "Year",
    y = "Number of Births",
    color = "Name",
    caption = "Source: Social Security Administration"
  ) +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 8)) +
  scale_color_manual(values = c("Javier" = "#2c3e50", "Micheal" = "#e67e22")) +
  theme_minimal(base_size = 13)

ggplotly(p_comp, tooltip = "text")
```

- **Micheal** peaked in **1957** with 4,831 births alongside the mid-century surge of the name Michael, but has since steadily declined.
- **Javier** surpassed Micheal around 1987 and has remained more popular ever since.
- In **2025**, there were **1,252** boys named Javier compared to **130** named Micheal.
